\chapter{Display String Tags}
\label{cha:display-string-tags}

This appendix lists all display string tags.
\begin{longtable}{|p{6cm}|p{8cm}|}
\hline
\tbf{p}[0] - x
&
X position of the center of the icon/shape; defaults to automatic graph layouting
\\ 
 \hline
\tbf{p}[1] - y
&
Y position of the center of the icon/shape; defaults to automatic graph layouting
\\ 
 \hline
\tbf{p}[2] - arrangement
&
Arrangement of submodule vectors. Values: row (r), column (c), matrix (m), ring (ri), exact (x)
\\ 
 \hline
\tbf{p}[3] - arr. par1
&
Depends on arrangement: matrix => ncols, ring => rx, exact => dx, row => dx, column => dy
\\ 
 \hline
\tbf{p}[4] - arr. par2
&
Depends on arrangement: matrix => dx, ring => ry, exact => dy
\\ 
 \hline
\tbf{p}[5] - arr. par3
&
Depends on arrangement: matrix => dy
\\ 
 \hline
\tbf{b}[0] - width
&
Width of object. Default: 40
\\ 
 \hline
\tbf{b}[1] - height
&
Height of object. Default: 24
\\ 
 \hline
\tbf{b}[2] - shape
&
Shape of object. Values: rectangle (rect), oval (oval). Default: rect
\\ 
 \hline
\tbf{b}[3] - fill color
&
Fill color of the object (colorname or \#RRGGBB or @HHSSBB). Default: #8080ff
\\ 
 \hline
\tbf{b}[4] - border color
&
Border color of the object (colorname or \#RRGGBB or @HHSSBB). Default: black
\\ 
 \hline
\tbf{b}[5] - border width
&
Border width of the object. Default: 2
\\ 
 \hline
\tbf{i}[0] - icon
&
An icon representing the object
\\ 
 \hline
\tbf{i}[1] - icon color
&
A color to colorize the icon (colorname or \#RRGGBB or @HHSSBB)
\\ 
 \hline
\tbf{i}[2] - icon colorization %
&
Amount of colorization in percent. Default: 30
\\ 
 \hline
\tbf{is}[0] - icon size
&
The size of the image. Values: very small (vs), small (s), normal (n), large (l), very large (vl)
\\ 
 \hline
\tbf{i2}[0] - overlay icon
&
An icon added to the upper right corner of the original image
\\ 
 \hline
\tbf{i2}[1] - overlay icon color
&
A color to colorize the overlay icon (colorname or \#RRGGBB or @HHSSBB)
\\ 
 \hline
\tbf{i2}[2] - overlay icon colorization %
&
Amount of colorization in percent. Default: 30
\\ 
 \hline
\tbf{r}[0] - range
&
Radius of the range indicator
\\ 
 \hline
\tbf{r}[1] - range fill color
&
Fill color of the range indicator (colorname or \#RRGGBB or @HHSSBB)
\\ 
 \hline
\tbf{r}[2] - range border color
&
Border color of the range indicator (colorname or \#RRGGBB or @HHSSBB). Default: black
\\ 
 \hline
\tbf{r}[3] - range border width
&
Border width of the range indicator. Default: 1
\\ 
 \hline
\tbf{q}[0] - queue object
&
Displays the length of the named queue object
\\ 
 \hline
\tbf{t}[0] - text
&
Additional text to display
\\ 
 \hline
\tbf{t}[1] - text position
&
Position of the text. Values: left (l), right (r), top (t). Default: t
\\ 
 \hline
\tbf{t}[2] - text color
&
Color of the displayed text (colorname or \#RRGGBB or @HHSSBB). Default: blue
\\ 
 \hline
\tbf{tt}[0] - tooltip
&
Tooltip to be displayed over the object
\\ 
 \hline
\tbf{bgp}[0] - bg x
&
Module background horizontal offset
\\ 
 \hline
\tbf{bgp}[1] - bg y
&
Module background vertical offset
\\ 
 \hline
\tbf{bgb}[0] - bg width
&
Width of the module background rectangle
\\ 
 \hline
\tbf{bgb}[1] - bg height
&
Height of the module background rectangle
\\ 
 \hline
\tbf{bgb}[2] - bg fill color
&
Background fill color (colorname or \#RRGGBB or @HHSSBB). Default: grey82
\\ 
 \hline
\tbf{bgb}[3] - bg border color
&
Border color of the module background rectangle (colorname or \#RRGGBB or @HHSSBB). Default: black
\\ 
 \hline
\tbf{bgb}[4] - bg border width
&
Border width of the module background rectangle. Default: 2
\\ 
 \hline
\tbf{bgtt}[0] - bg tooltip
&
Tooltip to be displayed over the module's background
\\ 
 \hline
\tbf{bgi}[0] - bg image
&
An image to be displayed as a module background
\\ 
 \hline
\tbf{bgi}[1] - bg image mode
&
How to arrange the module's background image. Values: fix (f), tile (t), stretch (s), center (c). Default: fixed
\\ 
 \hline
\tbf{bgg}[0] - grid tick distance
&
Distance between two major ticks measured in units
\\ 
 \hline
\tbf{bgg}[1] - grid minor ticks
&
Minor ticks per major ticks. Default: 1
\\ 
 \hline
\tbf{bgg}[2] - grid color
&
Color of the grid lines (colorname or \#RRGGBB or @HHSSBB). Default: grey
\\ 
 \hline
\tbf{bgl}[0] - layout seed
&
Seed value for layout algorithm
\\ 
 \hline
\tbf{bgl}[1] - layout algorithm
&
Algorithm for child layouting
\\ 
 \hline
\tbf{bgs}[0] - pixels per unit
&
Number of pixels per distance unit
\\ 
 \hline
\tbf{bgs}[1] - unit name
&
Name of distance unit
\\ 
 \hline
\tbf{ls}[0] - line color
&
Connection color (colorname or \#RRGGBB or @HHSSBB). Default: black
\\ 
 \hline
\tbf{ls}[1] - line width
&
Connection line width. Default: 1
\\ 
 \hline
\tbf{ls}[2] - line style
&
Connection line style. Values: solid (s), dotted (d), dashed (da). Default: solid
\\ 
 \hline
\end{longtable}
%%% Local Variables:
%%% mode: latex
%%% TeX-master: "usman"
%%% End:
