\chapter*{Document History}


\begin{longtable}{|l|p{1cm}|p{10cm}|}
\hline
\tabheadcol
\tbf{Date} & \tbf{Author} & \tbf{Change}\\\hline
%%
2005/07 & AV & updated for the {\opp} 3.2 release\\\hline
%%
2005/03 & AV & updated for the {\opp} 3.1 release\\\hline
%%
2004/12 & AV & updated for the {\opp} 3.0 release\\\hline
%%
2003/06 & AV & Mentioned Grace and ROOT in section "Visualizing...".
               Added section "Using STL in message classes".
               \\\hline
%%
2003/06 & AV & OMNeT++ 2.3 released\\\hline
%%
2003/04-06 & AV & "Design of OMNeT++" chapter revised, extended, and renamed to
               "Customization and Embedding".
               Added "Interpreting Cmdenv output" section to the
               "Running the Simulation" chapter. Added section about Akaroa
               in "Running the Simulation" chapter. Expanded section about
               writing shell scripts to control the simulation.
               Added background info about RNGs and warning about old RNG
               in "Class Library" chapter;
               revised/extended "Deriving new classes" section in same chapter.
               Bibliography converted to Bibtex, expanded and cleaned up;
               citations added to text.
               "Parallel Simulation" chapter: contents removed until new PDES
               implementation gets released.
               Revised and reorganized NED chapter. Section about message
               sending/receiving and other simple module related functions
               moved to chapter "Simple Modules"; cMessage treatment from
               "Simulation Library" merged with message subclassing chapter
               into new chapter "Messages". Deprecated cPacket.
               Removed sections "Simulation techniques" and
               "Coding conventions", and their useful fragments were incorporated
               elsewhere. Added/created sections about message transmission
               modeling, and using global variables. Added sections explaining
               how to implement broadcasts and retransmissions. Revised section
               about dynamic module creation. Deprecated putaside-queue,
               receiveNew(), receiveOn().
               Added section "Object ownership management"; removed section
               on "Using shared objects". \\\hline
%%
2003/03 & AV & OMNeT++ 2.3b2 released\\\hline
%%
2003/02 & AV & OMNeT++ 2.3b1 released\\\hline
%%
2003/01 & AV & Added chapter about message subclassing; revised chapter about
               running the simulation and incorporated new Cmdenv options; added new
               distributions and clarified many details in NED expr. handling section\\\hline
%%
Summer 2002 & Ulrich Kaage & Converted from Word to LaTeX\\\hline
%%
2002/03/18 & AV & Documented new ini file options about Envir plugins\\\hline
%%
2002/01/24 & AV & Refinements on the Parsec chapter\\\hline
%%
2001/10/23 & AV & Updated to reflect changes since 2.1 release (see include/ChangeLog)\\\hline
%%
\end{longtable}




%%% Local Variables:
%%% mode: latex
%%% TeX-master: "usman"
%%% End:
