
\chapter{References}
\label{cha:references}

\textbf{Simulation-related}


[JAIN91]\tab Jain, Raj: \textit{The Art of Computer Systems Performance
Analysis}. Wiley, New York, 1991.

[BFS86]\tab Bratley P., Fox, B. L. and Schrage, L. E.: \textit{A Guide
to Simulation}. Springer-Verlag, New York, 1986.

[JCH85]\tab Jain, Raj and Chlamtac, Imrich: \textit{The P}$^{\mathit{2}}$ \textit{Algorithm
for Dynamic Calculation of Quantiles and Histograms without Storing
Observations}, Communications of the ACM, 28(10), 1076-1085, 1985.

[PON91]\tab Pongor, Gy\"{o}rgy: \textit{OMNET: An Object-Oriented Network
Simulator}. 1991

[PON92]\tab Pongor, Gy\"{o}rgy: \textit{Statistical Synchronization: A Different
Approach of Parallel Discrete Event Simulation}. Lappeenranta
University of Technology, Data Communications Laboratory, Lappeenranta,
Finland, 1992

[PON93]\tab Pongor, Gy\"{o}rgy: \textit{On the Efficiency of the Statistical
Synchronization Method}. European Simulation Symposium (ESS'93),
Delft, The Netherlands, Oct. 25-28, 1993

[KOF95]\tab Kofoed, Stig: \textit{Portable Multitasking in C++}. Dr. Dobb's
Journal, November 1995. \href{ftp://ftp.mv.com/pub/ddj/1995/1995.11/mtask.zip}{ftp://ftp.mv.com/pub/ddj/1995/1995.11/mtask.zip}

%%
%% TBD update! include papers of Gabor Lencse
%%

\textbf{{\opp}-related research papers}


[VAR99]\tab \textit{''Using the {\opp} Discrete Event Simulation
System in Education''}. Andr\'{a}s Varga. IEEE Transactions
on Education, November 1999 CD-ROM issue; abstract in vol. 42,
no. 4, pp. 372, November 1999.

[VAR98a]\tab \textit{''K-split - On-Line Density Estimation for
Simulation Result Collection''}. Andr\'{a}s Varga. In the
Proceedings of the European Simulation Symposium (ESS'98). October
26-28, 1998. Nottingham, UK.

[VAR98b]\tab \textit{''Parameterized Topologies for Simulation Programs''}.
Andr\'{a}s Varga. In the Proceedings of the Western Multiconference
on Simulation (WMC'98) / Communication Networks and Distributed
Systems (CNDS'98). January 11-14, 1998. San Diego, CA.

[V\&F97]\tab \textit{''The K-Split Algorithm for the PDF Approximation
of Multi-Dimensional Empirical Distributions without Storing
Observations''}. Andr\'{a}s Varga and Babak Fakhamzadeh. In
Proceedings of the 9th European Simulation Symposium (ESS'97),
pp.94-98. October 19-22 1997, Passau, Germany.

[V\&P97]\tab \textit{''Flexible Topology Description Language for
Simulation Programs''}. Andr\'{a}s Varga and Gy\"{o}rgy Pongor.
In Proceedings of the 9th European Simulation Symposium (ESS'97),
pp.225-229. October 19-22 1997, Passau, Germany.


\textbf{Former {\opp} documents}


[OMN1]\tab Vass Zolt\'{a}n.: \textit{PVM Extension of {\opp} to Support
Statistical Synchronization}. Diploma Thesis, Technical University
of Budapest, 1996 (in Hungarian).

[OMN2]\tab Andr\'{e} Maurits, George van Montfort and Gerard van de
Weerd: \textit{{\opp} extensions and examples}. Technical University
of Budapest, Dept. of Telecommunications, 1995.

[OMN3]\tab Jan Heijmans, Alex Paalvast, Robert van der Leij: \textit{Network
simulation using the JAR compiler for the {\opp} simulation
system}. Technical University of Budapest, Dept. of Telecommunications,
1995.

[OMN4]\tab Varga Andr\'{a}s.: \textit{{\opp} - Portable User Interface
for the {\opp} Simulation System}. Diploma Thesis, Technical
University of Budapest, 1994 (in Hungarian).

[OMN5]\tab Lencse G\'{a}bor: \textit{Graphical Network Editor for {\opp}}.
Diploma Thesis, Technical University of Budapest, 1994 (in Hungarian).

[OMN6]\tab Varga Andr\'{a}s.: \textit{{\opp} - Portable Simulation Environment
in C++}. TDK work, Technical University of Budapest, 1992 (in Hungarian).


\textbf{Other simulation software}


See web site


\textbf{C++ language}


Too many books to list.


\textbf{Cygwin}


[CYGWIN]\tab \href{http://sourceware.cygnus.com/cygwin/top.html}{http://sourceware.cygnus.com/cygwin/top.html}

%%
%%\textbf{DJGPP}
%%
%%
%%[DJGPP1]\tab Official DJGPP Home Page: \href{http://www.delorie.com/djgpp}{http://www.delorie.com/djgpp}
%%


\textbf{PVM}

[PVM1]\tab The Official PVM Home Page. \href{http://www.epm.ornl.gov/pvm/pvm\_home.html}{http://www.epm.ornl.gov/pvm/pvm\_home.html}

[PVM2]\tab \href{http://www.sp2.uni-c.dk/PVM/PvmIntro.html}{http://www.sp2.uni-c.dk/PVM/PvmIntro.html}

[PVM3]\tab \href{http://www.cse.ogi.edu/DISC/projects/mist/related-work/pvm.html}{http://www.cse.ogi.edu/DISC/projects/mist/related-work/pvm.html}

%%
%% textbf{Turbo Vision}
%%
%% [TV1]\tab \textit{Borland C++ 3.1 Manuals}. Borland International, 1992.\newline
%% [TV2]\tab The TVPlus Archieve. \href{http://wvnvm.wvnet.edu/~u6ed4/tvhome.htm}{http://wvnvm.wvnet.edu/\ensuremath{\sim}u6ed4/tvhome.htm}\newline
%% [TV3]\tab Sierwald, Joern: 32-bit Portable Turbo Vision. \href{http://wvnvm.wvnet.edu/~u6ed4/tvptsier.htm}{http://wvnvm.wvnet.edu/\ensuremath{\sim}u6ed4/tvptsier.htm}
%%


\textbf{Tcl/Tk}

[TCLTK1]\tab Welch, Brent: \textit{Practical Programming in Tcl and Tk}. Prentice-Hall, 1995

%%
%% [TCLTK2]\tab HyperTcl. \href{http://web.cs.ualberta.ca/~wade/HyperTcl/}{http://web.cs.ualberta.ca/\ensuremath{\sim}wade/HyperTcl/}
%%
%% [TCLTK3]\tab TCL WWW Info. \href{http://www.sco.com/Technology/tcl/Tcl.html}{http://www.sco.com/Technology/tcl/Tcl.html}
%%


\textbf{Gnuplot}


[GPLOT1]\tab Brief tutorial:\\
\href{http://nacphy.physics.orst.edu/DATAVIS/datavis.html}{http://nacphy.physics.orst.edu/DATAVIS/datavis.html}


[GPLOT2]\tab Reference:\\
\href{http://www.cm.cf.ac.uk/Latex/Gnuplot/gnuplot.html}{http://www.cm.cf.ac.uk/Latex/Gnuplot/gnuplot.html}


[PMTV1]\tab PlotMTV:\\
\href{http://cauchy.math.edu/workshop/Plotmtv/plotmtv.html}{http://cauchy.math.edu/workshop/Plotmtv/plotmtv.html}


\textbf{Xmgr}

[XMGR1]\tab Brief tutorial:\\
\href{http://nacphy.physics.orst.edu/DATAVIS/xmgr.html}{http://nacphy.physics.orst.edu/DATAVIS/xmgr.html}


%%% Local Variables:
%%% mode: latex
%%% TeX-master: "usman"
%%% End:
