\documentclass[a4paper]{article}
\usepackage{url}
\usepackage{fullpage}
\usepackage{bbold}

\date{\today}
\author{Nicky van Foreest}

\title{Simulating Queueing Networks with OMNeT++}

\begin{document}
\maketitle \abstract{\noindent This documents provides a quick
  introduction to OMNeT++, with a focus on queueing applications. The
  aim is to get the interested user up and running with a simple M/M/1
  fifo queueing demo. After this the user can start exploiting more of
  OMNeT++'s functionality based on the extensive user manual and the
  sample directories.}
\tableofcontents

\section{Introduction}
\label{sec:introduction}
OMNeT++ is a discrete event simulation environment based on C++. Andr{\'a}s
Varga is the principal author and maintainer.

When I started to use it for simulating queueing networks, I was
somewhat awed by the amount of functionality that OMNeT++ provides and
the size of the user manual. Besides this, I had no experience with
C++ programming, which was an extra hurdle in becoming acquainted with
the power of the tool. Hence I had to invest some time to find out
which parts and functionalities of OMNeT++ are specially useful for
queueing systems analysis, and how to get these to work. I wrote this
tutorial (this document together with some code examples) with the
intention that a  user new to OMNeT++ has a simple starting point from
where he/she can further explore this simulation tool.

In this tutorial I assume that the reader has experience with
programming, at least with C, and has some (very) basic understanding
of queueing theory, say the first few sections of the chapter on
queueing theory of~\cite{Ross93}.  Furthermore, I will not explain the
details of the code. These are for the reader to study. This may seem
like a thread, but the code is not hard to understand.

Perhaps I should have started by mentioning the advantages of using
OMNeT++. I refrain from doing this at this point of the document, but
instead postpone it to the end. The reason for this is twofold. In the
first place the OMNeT++ user manual and the OMNeT++ home page mention
plenty of examples. In the second place, I want to focus on my
personal experiences, and this is less important than having you start
playing with OMNeT++ and this demo.

\subsection{Where to Get the Software?}
\label{sec:where-get-software}
You can find OMNeT++ at \url{http://whale.hit.bme.hu/omnetpp/} If you
are too lazy to type this in, a search with \url{google.com} on
\texttt{omnetpp} will give an instant hit. On my homepage
\url{http://www.math.utwente.nl/~foreest/} you can find the latest
release of the queueing software, a library of basic queueing systems,
such as a multi-server queue, a processor sharing queue, etc.  If you
want to make advanced plots of the simulation results, make sure to
get gnuplot at \url{http://www.gnuplot.info}

\subsection{Feedback}
\label{sec:feedback}
Please do not hesitate to send me your comments, especially if you
think the code or this document contains inaccuracies, unclear
passages, etc. When you have something to contribute to the software
library, please inform me so that I can include it in this package. I
grant full acknowledgment with respect to authorship.

My email address is: \url{n.d.vanforeest@math.utwente.nl}, or
\url{nokfi@hotmail.com}.

OMNeT++ related questions can be posted at the OMNeT++ mailing list:
\url{omnetpp-l@it.swin.edu.au}. Chances are high that Andr{\'a}s himself
answers your questions.


\subsection{Structure of This Document}
\label{sec:struct-this-docum}
In Chapter~\ref{sec:simulating-mm1-fifo} I discuss a working example
of a queueing simulation in OMNeT++.  The implemented example is the
archetypal M/M/1 fifo queue.  Chapter~\ref{sec:next-steps} shows some
methods to change the working of the simulation, the simulation
parameters, the simulation output, and the like.
Chapter~\ref{sec:impl-fifo-simul} discusses the
implementation of the M/M/1 example.

\subsection{Acknowledgments}
\label{sec:acknowledgements}
\begin{itemize}
\item Andr{\'a}s: Thanks a lot for OMNeT++.
\item Phillip M. Feldman pointed out a few typos and added footnote~\ref{fn:1}.

\end{itemize}

\subsection{Versions}
\begin{enumerate}
\item[0.1] January 2001
\item[1.0] May 2001
\item[2.0] June 2002
\item[2.1] October 2002
\item[2.2] January 2003
\end{enumerate}

\section{Simulating the M/M/1 Fifo Queue}
\label{sec:simulating-mm1-fifo}


This chapter focusses on how to set up, run, and obtain output of the
M/M/1 simulation in OMNeT++. Since I want to show as many features of
OMNeT++ as possible,  this example is a bit baroque in its output.


\subsection{Getting the Simulation to Run}
\label{sec:getting-it-running}
Before running the M/M/1 example it is a good idea to check that
OMNeT++ and all the software it depends on is installed properly. I
usually test this by running one of the standard samples included in
the OMNeT++ distribution, such as the Nim game in
\texttt{samples/nim/}.

Once you are convinced that OMNeT++ works, run \texttt{opp\_makemake
  -f} in the directory that contains the source files of the queueing
tutorial, supposedly \texttt{tutorial/queuesTut/}, to make a
\texttt{Makefile}. The binary \textbf{opp\_makemake} should be
provided with OMNeT++.  In case it cannot be found, try running it
directly from the \texttt{src/utils/} directory of the OMNeT++
distribution. Once you have a \texttt{Makefile}, a simple
\texttt{make} should do to get the binaries for the fifo simulation.

A general remark is of importance here. In case you change something
in one of the files, you should run \texttt{make} again. When you
decide to copy all files to another directory, or include other files
in this directory, run \texttt{opp\_makemake -f } again. Run
\texttt{make clean} to remove all object files and binaries.

The binary to run has the same name as the directory containing the
Makefile that was generated by \texttt{opp\_makemake}. In my case this
is \texttt{queuesTut}. Run it, i.e., type \texttt{./queuesTut} at the
prompt. If everything works the way it should, a few windows should
pop up.

I consider it a bit too much work to type in \texttt{make} and then
the binary  again and again. Therefore I included a very
simple script \texttt{run} which does this for me.

As an aside, the fifo simulation that is part of the tutorial is not
the same as  \texttt{fifo1} and \texttt{fifo2} of the \texttt{samples}
directory. The one provided in the tutorial is a stripped down version
that includes just the bare minimum to simulate a fifo queue. The
\texttt{fifo1} and \texttt{fifo2} show some more advanced tricks with OMNeT++.

\subsection{Running the Demo}
\label{sec:running-simulation}
Once the simulation has started you should see two windows:
\begin{itemize}
\item A graphics window showing a small queueing network: a job
  generator, a fifo queue, and a job sink.
\item The navigation window, called \texttt{OMNeT++/TKenv}, containing
  the main functionality to run the simulation.
\end{itemize}

Double click on the fifo widget on the canvas of the graphics window.
A new box should appear, called \texttt{(Fifo) fifonet.fifo[0]}. Click
on \texttt{Objects/Watches}.  You will then see a few lines with
text.. The line with the string \texttt{(cQueue)} contains the word
\texttt{(empty)}.  This will change to the number of customers in
queue (= one less the number in the system when a customer is served)
once the simulation is running.  Then double click on the line with
the word \texttt{Job Distribution}.  You will see a blank blue (on my
screen, that is) window. Here a histogram will show, once the
simulation started, the probability density function of the number of
jobs in the system as seen by an arriving job. Finally, click the line
\texttt{Jobs in system}. A yellow canvas will come up. This will
contain a simple plot of the number of jobs in the system, the one in
service included.

Now you can start the simulation by pressing the \texttt{RUN}-button
in the \texttt{OMNeT++/TKenv} window. In the graphics window you
should see jobs hopping from the generator to the queue, and from the
queue to the sink.  Press \texttt{STOP} after several seconds. The
first line of the \texttt{(Fifo) fifonet.fifo[0]} window will now,
with probability $ > 0$, show that the queue contains some customers.
When you double click on this line, another window will tell you which
jobs are currently waiting in the queue.

To make a distinction between individual customers may seem not very
interesting from a queueing perspective; the M/M/1 queueing model
assumes that customers do not have real identity---they only differ by
their service requirements and arrival time. However, when studying
more complicated models, for instance when jobs have different type,
this information becomes relevant. It becomes even more important if
you want to study protocol and queue interactions, such as when
multiple TCP sources share buffered resources, i.e., routers.

The histogram window should now contain a few black bars. When you put
your mouse pointer on one of these, it will change color. (Mine
becomes gray.) There is a line at the bottom of window that shows
the number of arrivals that observed a certain number of jobs in front
of it in the system.  In other words, if cell \# 1 contains 10, this
means that 10 arrivals saw one job in the
system\footnote{\label{fn:1}When the arrival process is Poisson, i.e.,
  the interarrival time is exponentially distributed, the distribution
  of the number of customers seen by arrivals is the same as the time
  average of the number of customers in the queue. This phenomenon is
  due to the PASTA (Poisson Arrivals See Time Averages) property, see
  for instance \cite{Ross93} or
  \cite{el-taha98:_sampl_path_analy_queuein_system}. In general you
  should be careful to chose a sampling method of the simulation that
  provides the statistics you want to collect. Beware!}.

Clicking  the \texttt{Job Distribution} line in the
\texttt{(Fifo) fifonet.fifo[0]}-window with the right mouse button, enables
you to choose between a graphical representation of the gathered data,
and a textual one, called \texttt{Object}. I discussed the graphical
one above. The text information will provide you with aggregate statistics
such as the mean number of jobs found upon arrival, etc.

The last step will be to press \texttt{EXPRESS} in the
\texttt{OMNeT++/TKenv} window; I leave the other
buttons for you to discover. In fact, my best general advice is to
press on any buttons you may see, and find out what it does. Do not
forget that often both mouse buttons, the left and the right, can be
used for different effects.


\subsection{Analyzing the Simulation Results}
\label{sec:interpreting-results}
If you have not interrupted the simulation by pressing the big red
\texttt{STOP} button, the simulation will finish after having
generated 5010 jobs. On my machine, which is not particularly fast,
this takes less than a second.

Notice that the textual and graphical representations of the histogram
are updated after the simulation has finished. The
\texttt{OMNeT++/Tkenv} window contains as well a number of results.
You might need to scroll a bit up and down in the window to view all
the simulation results.

The directory will now contain a few new files as well.
\begin{itemize}
\item \texttt{fifo.sca} This file contains the statistics that were
  gathered in the variable \texttt{jobDist} of class
  \texttt{cDoubleHistogram}, see the user manual section 6.13, such as
  the number of jobs generated, the histogram data, etc. Please have a
  look at it now, which can be done by pressing \texttt{view->Output
    Scalar File} in the taskbar of the main window. (For some reason,
  the file \texttt{fifo.sca} remains empty after the simulation
  finished. One trick to resolve this, is by pressing \texttt{Step}
  once again. OMNeT++ will ask you to rerun the network. Say
  \texttt{yes}. Then the content of \texttt{fifo.sca} will appear.)
  The textual info of the histogram should be contained in this file
  too.
\item \texttt{fifo.vec} This file contains info about the dynamics of
  the number of jobs in the system. You should process this with the
  OMNeT++ tool \texttt{plove}. Chapter 9 of the user manual contains
  the instructions.  Briefly, start up \texttt{plove}. The left most
  button allows you to \texttt{load} the file. Send it to the right
  window with the arrow button in the middle. Press \texttt{PLOT!}.
  Now \texttt{gnuplot} should fire up and show the dynamics of the job
  distribution. You may have to fiddle around with the
  \texttt{options} to get a nice graph.
\end{itemize}

In general you can process these files, with \texttt{awk} for instance,
or another C++ program for that matter, to postprocess it. You might
instead want to do the post-processing  in the function
\texttt{Fifo::finish}, about which I will talk later. Here comes in
the some of the power of OMNeT++'s use of   a real programming language:
you can incorporate your own post-processing functions in the
simulation itself, in case you want to.

\paragraph{}
Summarizing, OMNeT++ provides a number of ways to present simulation
data:
\begin{itemize}
\item dynamically, by means of jobs jumping from one node in the
  network to another;
\item graphically, by means of dedicated collection classes such as
  \texttt{cOutVector};
\item verbally, by means of the \texttt{ev <<} statement in the code
  which is output to the  canvas of the \texttt{OMNeT++/Tkenv} window;
\item by means of files.
\end{itemize}

\section{Modifying the Basic Example}
\label{sec:next-steps}
The previous chapter discussed only one type of queue, the M/M/1
queue, with one specific parameter setting. You probably want to
simulate other types of queues, more generic service distribution,
etc. Another interesting degree of freedom is to speed up the
simulation by leaving out the windows.

\subsection{Changing the  Simulation}
\label{sec:changing-basic-mm1}
I will discuss some ways to change and extend the current simulation
environment. The most important way to do this is with the
\texttt{omnetpp.ini} file, see chapter 8 of the user manual for all
details.

\paragraph{Changing the Interarrival and Service Rate}
\label{sec:chang-inter-serv}
Change the numbers in \texttt{omnetpp.ini} related to the exponential
distribution. Mind that the interarrival rate and service rate are the
inverse of the parameters you specify, e.g.
\texttt{fifonet.fifo[1].service\_time = exponential(2)} means that the
expected service duration $ = 1/\mu = 2$.

Besides with \texttt{omnetpp.ini} file, you can change these values by
pressing \texttt{Params} in the \texttt{(Fifo) fifonet.fifo[0]} window.
Click for instance on the line with \texttt{service\_time}. Now you
edit the field in the window that will appear. To try, for instance,
another distribution, type in its name and parameters, see the next
paragraph for more on this.  Do not forget to press the \texttt{Enter}
key on your keypad to make the edit effective.

\paragraph{Changing the Interarrival and Service Distribution}
\label{sec:chang-inter-serv-1}
OMNeT++ provides a few standard distributions, such as the uniform and
exponential distribution, see sections 4.9.5 and 6.13 of the user
manual. In case you need other distributions, you have to build them
yourself. This is in itself quite interesting, see Ross~\cite{Ross93},
and, of course, Knuth\cite{Knuth97}. Once you know mathematically
speaking what to do, OMNeT++ makes the implementation very easy.  You
define your new distribution in the file \texttt{distributions.cc}. I
have already provided a few simple examples.  Note that you can use
the standard distributions of OMNeT++ in your new functions right
away. Once you have implemented your new distribution, you can simply
use its name in \texttt{omnetpp.ini} and pass appropriate parameters
to it. As an example, if you would like to use the distribution
\texttt{perturbedExponential}, which you can find in the file
\texttt{distribution.cc}, as service distribution, than you should use
the following line in \texttt{omnetpp.ini}:
\texttt{fifonet.fifo[0].service\_time = perturbedExponential(2,1)}

\paragraph{Changing the Network}
\label{sec:changing-network}
The \texttt{omnetpp.ini} parameter \texttt{fifonet.num\_buffers}
enables you to make a network of tandem queues. Changing it from the
current value 1 to $k$, say, will put $k$ queues in tandem. You should
as well change the line \texttt{fifonet.fifo[0].service\_time =
  exponential(2)} to \texttt{fifonet.fifo[*].service\_time =
  exponential(2)}, that is, the \texttt{0} has to change to \texttt{*}
to reference all fifos, instead of just the first one with id
\texttt{0}. I actually chose a slightly different approach in
\texttt{omnetpp.ini}. First I give \texttt{fifo[0]} its value, and
then the others by means of the \texttt{*}.

You will find included as well a ring network. This is built out of
simple fifo queues that are connected sequentially to each other. If
you want to run it, you have to change a few lines in
\texttt{omnetpp.ini}. This file contains a few remarks on where the
changes should be made. Since the ring is a closed queueing network,
there are no external arrivals of jobs nor sinks. Therefore these
two modules are not included in the ring. Furthermore, the buffers
should contain some initial number of jobs that will start circling
around. You can set these numbers by the parameter
\texttt{ring.fifo[*].num\_init\_jobs = 20} to be found in
\texttt{omnetpp.ini}.\footnote{For the interested. If you analyze the
  expected number of jobs in the queues, and add them, it will appear
  as if one job is missing, this is the application of the Arrival
  Theorem, see e.g.~\cite{Ross93}.}

More general networks are for you to build. Section 4.10 and 6.19 of
the manual will tell you how. You should as well consult the following
OMNeT++ samples. The sample directory \texttt{token} provided with the
standard OMNeT++ distribution shows a circular network. The
\texttt{fddi} sample shows a large network. (Do not forget to click on
one of the rings in the \texttt{fddi} sample to see how complicated
a network you can actually simulate with OMNeT++.)


\paragraph{Changing the Job Scheduling}
\label{sec:chang-job-sched}
I have implemented, at the moment, some other queueing models. In case you are
interested, please have a look at the queues package, to be found on
my homepage or the OMNeT++ server.

\paragraph{Changing the random number generator}
\label{sec:chang-rand-numb}
For the more suspicious of you, section 6.9 of the OMNeT++
user manual discusses the random number generator. You can replace
this with your own if you want.


\subsection{Speeding up the Simulation}
\label{sec:speed-up-simul}
If you are convinced that everything works the way it should, and you
are just interested in numerical output, you can run the simulation
straight from the prompt with the \texttt{cmdenv} mode. No more
windows will appear, only the output files will be produced. Due to
this, the simulation will become quicker as well.

To achieve this, change these lines in the  \texttt{Makefile}

\begin{verbatim}
# User interface (uncomment one) (-u option)
#USERIF\_LIBS=\$(CMDENV\_LIBS)
USERIF\_LIBS=\$(TKENV\_LIBS)
\end{verbatim}

to

\begin{verbatim}
# User interface (uncomment one) (-u option)
USERIF\_LIBS=\$(CMDENV\_LIBS)
#USERIF\_LIBS=\$(TKENV\_LIBS)
\end{verbatim}

Do a \texttt{make clean} and \texttt{make} to remove all window
related code from the simulation executable. Now it should work. Be
aware that a new \texttt{opp\_makemake -f} reverts the makefile to the old
situation, i.e., the simulation with Tk windows.

You will want these options in the following section of
\texttt{omnetpp.ini}:
\begin{verbatim}
[Cmdenv]
runs-to-execute = 1
express-mode = yes
# for non-express mode:
module-messages = yes
event-banners = yes
# for express mode:
status-frequency = 50000
performance-display = no
\end{verbatim}
Play with these options to discover that the number of lines of
simulation output will be a bit too much to handle. See section 8.5 of
the manual for more info.

The user manual contains as well some more hints to speed up the
simulation still further, see sections 6.18 and  8.7.

\section{The Implementation of the M/M/1 Simulation}
\label{sec:impl-fifo-simul}
Before starting the main subject of this chapter, I need to define one
concept: a \emph{functional entity}. A functional entity is a part of
a simulation that carries out a specific action on a job, or a
message. For example, a server or a message generator are functional
entities. They are, so to say, the essential functional units that take
care of one process step in the lifetime of a job, or message. The
implementation of such entities will be called \emph{modules}. Now
back to the simulator.

The simulation environment is built out of, mainly, two types of
files: \texttt{.ned} files and \texttt{.cc} files.  The \texttt{.ned}
files roughly describe how the functional entities connect to each
other and communicate; the \texttt{.cc} files contain the C++ code
that implement the behavior of the entities. These file types I will
discuss in some more detail below. I expect you to have the
\texttt{.ned} and \texttt{.cc} files belonging to this demo at hand.

After having read this, be sure to check out the discussion of the Nim
game in the manual as well, and the samples \texttt{fifo1} and
\texttt{fifo2}.  They are instructive and show additional
functionality of OMNeT++ relevant for queueing systems analysis.

\subsection{A Simple Method to Understanding the Code}
There are various ways to  understand the implementation of a
new simulation. I found the following approach the most useful. First I
run it, of course, to get an understanding of where the various
entities reside and how they exchange jobs. Then I work through the
files belonging to each entity separately. I start with the
\texttt{.ned} file to understand what goes in and out of a module, and
the parameters it needs.  Then I give the header file a brief look to
become familiar with the module's specific internal variables and
functions.  Finally I study the C++ code belonging to the module.
Once I somewhat understand what this first module does, I take the
next module in the simulation chain, that is, the module that gets its
messages from the one I just studied. Working this way, I gain a
quicker understanding of what is going on than by first working
through all \texttt{.ned} files, than all the C++ files, etc.

In this document I will not, however, follow the above suggestion,
mainly for brevity. Here I only want to illustrate certain key points
of the simulation, and leave the studying to you.

\subsection{The \texttt{.ned} Files}
\label{sec:texttt.ned-files}
Each entity in a simulation needs to communicate via \emph{messages}
with itself and other entities. Messages can be used for various
purposes. One is to represent jobs that need service at queues.
Another is to convey information of the entity's state to itself in
the future, or to other entities.

Specifically, the fifo example contains three entities: a source module,
a fifo module and a sink module. The source module generates messages
that represent jobs.  These job-messages are sent to the fifo module.
This in turn delays the messages according to the present queue, then
services them, and finally sends the messages to the sink module. The
sink module processes these messages to extract some final statistics.
The sink is where the messages leave the queueing network.  The sink
module releases as well the memory allocated to the job messages.

Clearly, all these modules need in- and output gates to receive and
send messages. The \texttt{.ned} file specifies these gates. In our
example, the fifo module has an input and an output gate. Furthermore
these modules may need parameters such as service rate, queue size,
etc. These parameters need be \emph{declared}  in the \texttt{.ned} files
too. The \emph{definition} of the parameters, i.e., giving them a
value, takes place when the simulation starts.

The next step is to glue all the modules together to form a network.
This should be done by means of another \texttt{.ned} file that
defines a \emph{compound module}. This new module uses the simple
modules and connects them into networks. Besides this, it passes
parameter values on to them during the simulation. Finally it gives
some directives on where on the canvas the simple modules should
appear. To state this in terms of the fifo queue, \texttt{FifoNet},
defined in \texttt{fifonet.ned}, connects the output of the source to
the input of the fifo queue, and the output gate of the queue to the
sink. Consult \texttt{fifonet.ned} for all details. Take especially
notice of the syntax, and be aware of when to use a comma `,', and a
semicolon `;'. I made some time consuming errors by not using them in
the appropriate way.

Note that this hierarchy of
modules---simple modules forming compound modules, forming in turn
other compound modules--provides a very efficient way to build large,
complex queueing networks. Start by building the parts, and then
connect these parts to larger compound modules up to entire networks.

\subsection{The \texttt{.cc} and \texttt{.h} Files}
\label{sec:texttt.cc-texttt.h-f}
Now we have defined all entities that take part in the process cycle
of job-messages, and specified the routing between these entities in
the compound modules, we should tell how each entity should process
job-messages. In concrete terms, the generator should generate jobs
with certain interarrival times distributed according to some
specified probability distribution. These jobs are represented by a
message, so that in fact the generator generates messages. Then it
sends these messages out of its output gate. If you will have a look
at \texttt{gen.cc} you will see that it will produce in total
\texttt{fifonet.gen.num\_messages } messages, each separated some time
\texttt{ia\_time} apart. These two parameters are defined in the file
\texttt{omnetpp.ini}.

\texttt{fifo.cc} contains the heart of the simulation. Once you
understand this code, which may be a bit hard at first, you have
mastered the most important aspects of OMNeT++.

Once the jobs arrive at the fifo queue they will be queued or not,
depending on whether the server is idle or not.  These arriving jobs
are handled in the function \texttt{Fifo::handleMessage}. Lets see
what it does.  Suppose the server is busy. Then the event stack of the
simulator should contain an event that indicates when the job's
service is supposed to be finished.  If you think a bit about how to
implement such events, taking into account that all events are related
to messages, you will understand that the server should send itself an
\texttt{endService} message at the moment a job's service starts, to
indicate when this job's service should stop.  This
\texttt{endService} message will be put on the event stack by the
event scheduler (not to be confused with the queueing scheduler!). At
some point in time, the event scheduler will remove this message from
the stack, and will give it to the fifo module.  Once this module
receives the \texttt{endService} message, it knows that the job's
service has ended, and that it should send it to the sink. When the
queue is empty after the departure, the server should just wait. On
the other hand, if there are jobs in queue, it should take the job in
front of the queue, and start another service period. Now have a look
at \texttt{fifo.cc} and think deeply about it until you understand
what is exactly going on. Remember well that a simulation entity has
no other means to communicate with itself in the future than via
self-messages, and that these messages have to be put on the internal
event stack.

\texttt{fifo.cc} contains four more functions. The functionality of
\texttt{Fifo::initialize} should be clear from its implementation. The
function \texttt{Fifo::finish} processes part of the simulation
results when the simulation has finished.  The other two functions
\texttt{Fifo::serviceRequirement} and \texttt{endService} enable you
to specify what to do when a job starts service---here only a random
service duration is generated---and when it leaves. This provides
interesting flexibility with respect to queueing simulation. For
instance, in multi-class networks, jobs may change class from service
station to service station, or a fraction may be sent back for another
processing step at the server, etc. etc. Do not be tempted too easily
to handle this kind of functionality in the \texttt{handleMessage}
function. It can become quite complex.

The last module  of interest, the sink, should be a breeze once
you have mastered the above. It is a good idea to think about why we
compute the waiting time of a job is this module, and why this is not
possible in \texttt{fifo.cc}. This is of course by no means a generic
property of OMNeT++, but just a consequence of the current
implementation of the queueing entity.

\section{Why Use OMNeT++ for Queueing System Simulation?}
\label{sec:why-use-omnet++}
Some time ago I worked for a large telecommunication company. In one
of the projects in which I participated, we tried to reduce the
convergence time of a distributed network restoration protocol. The
interaction between the protocol, the involved timers, and the states
of the queues in the equipment became soon completely intractable,
analytically speaking.  Hence we decided to analyze the behavior of
the network and equipment with a simulator. As the developers wrote
the protocol in C++, we wanted to use this very code in a simulator.
Besides this we needed a graphical environment to see whether the
protocol messages traveled the way we wanted, and ended up in the
right queues. OMNeT++ provided this, and more.

The general experiences I gained during this and other projects with
respect to simulation complex systems were briefly as follows:
\begin{itemize}
\item Using production code in a simulator saves a major amount of
  time compared to having to redesign the work in some kind of
  `high-level', seemingly user friendly simulation environment.
  Besides this, the simulation itself helps to test the real code,
  that is, the code that will be actually  implemented in commercial
  equipment.
\item Having insight in the code of the simulator is
  helpful to understand details of the simulation. Furthermore, it can
  be necessary to extend parts of the tool's functionality. Being
  dependent on companies to give you support, but keeping you securely
  away from the details of the implementation of their simulator, is
  not always what you want. Worse yet, when you have a problem with the
  simulator you have to wait for their answer, which does not always
  relate to your question \ldots :-(.
\end{itemize}
Based on the above experiences as well as on some past work with
commercial simulators, I want to point out some advantages, at least
in my opinion, of running queueing simulations within OMNeT++.
\begin{itemize}
\item The behavior of  systems is programmed in a `normal'
  programming language, in this case C++. This provides the user way
  more flexibility than a `high-end', tool-specific programming
  environment. For instance, if you use Bones, your programming skills
  gained during working with Bones only apply to the Bones
  environment. Besides this, these high-end languages appear, as long
  as you stick to the provided tutorials, to be quite generic. But
  when you try your hands on your own examples,  you will feel how
  restrictive these tool-specific languages sometimes  are.
\item As a consequence of the use of C++, OMNeT++ is flexible and
  extensible. As an example, you can easily implement your own job
  interarrival and service distributions, statistical tests on the
  simulation results, more general networks, scheduling distributions,
  etc.
\item It is under GPL license.
\item OMNeT++ enforces `separation of concerns'. You have to specify
  the separate functional entities, such as queues, schedulers, job
  generators, separately from routing functionality, that is the
  definition of the network topology.
\end{itemize}
Let me say one more thing about the first bullet. Of course, these
commercial tools allow you to specify parts of the system in your own
code, and connect this code with the simulation platform by means of
software hooks.  But then you have to figure out how this works plus
all debugging, and you still have to program in a real programming
language.

With respect to the last item, I hesitated to use the
possibly vague words `object-oriented', but that is what it really is.
In my experience this is a bonus, as it makes the simulation
environment  modular. You can easily change only one part, or a
module. As long as the interfaces between the modules remain the same,
everything will remain working. Changing the topology is easy
too. Take for instance this demo. If you want to build Jackson
networks---a bunch of M/M/1 queues connected in a network such that
jobs can arrive and leave the network---you only have to figure out
how to set up such networks.  the functionality of job sources, sinks
and service stations is already there, ready to use. If you want to
change some of the service rates, you can easily do this in the
\texttt{omnetpp.ini} file.

Let me stop here, and let you convince yourself about the uses of
OMNeT++.

%\begin{thebibliography}{Ros93}

%\bibitem[Knu97]{Knuth97}
%D.E. Knuth.
%\newblock {\em The art of computer programming}, volume 2, Seminumerical
%  algorithms.
%\newblock 3 edition, 1997.

%\bibitem[Ros93]{Ross93}
%S.M. Ross.
%\newblock {\em Introduction to Probability Models}.
%\newblock Academic Press, 5th edition, 1993.

%\end{thebibliography}

\bibliographystyle{alpha}
\bibliography{biblio_nicky,mystuff}

\end{document}

%%% Local Variables:
%%% mode: latex
%%% mode: outline-minor
%%% TeX-master: t
%%% End:


